Supporting Information Appendix:\\ Black-swan events in animal populations

### 2017-01-13 text/ms/ms_pnas.tex
Resubmit

% \input{../analysis/values} % R output
\newcommand{\basePriorMean}{102}
\newcommand{\basePriorMedian}{71}
\newcommand{\basePriorProbHeavy}{7.7}
\newcommand{\medianTimeSteps}{26}
\newcommand{\meanTimeSteps}{30.2}
\newcommand{\minTimeSteps}{20}
\newcommand{\maxTimeSteps}{117}
\newcommand{\birdN}{191}
\newcommand{\insectsN}{182}
\newcommand{\mammalsN}{125}
\newcommand{\fishN}{108}
\newcommand{\birdNH}{14}
\newcommand{\insectsNH}{5}
\newcommand{\mammalsNH}{6}
\newcommand{\fishNH}{0}
\newcommand{\birdPH}{7}
\newcommand{\insectsPH}{3}
\newcommand{\mammalsPH}{5}
\newcommand{\fishPH}{0}
\newcommand{\NOrdersHeavy}{16}
\newcommand{\POrdersHeavy}{41}
\newcommand{\baseFiftyObsFiftySwitch}{8}
\newcommand{\baseSeventyFiveObsFiftySwitch}{2}
\newcommand{\totalHeavyFifty}{26}
\newcommand{\totalHeavySeventyFive}{17}
\newcommand{\baseFiftyObsFiftySwitchPerc}{31}
\newcommand{\baseSeventyFiveObsFiftySwitchPerc}{8}
\newcommand{\baseNuTenObsTenSwitch}{8}
\newcommand{\baseNuTen}{26}
\newcommand{\baseNuFiveObsTenSwitch}{2}
\newcommand{\pHeavyNThirty}{0.13}
\newcommand{\pHeavyNSixty}{0.20}
\newcommand{\pIncHeavyNThirtyNSixty}{1.6}
\newcommand{\obsErrorNuFivePerc}{75}
\newcommand{\modelsNoConvergeAROne}{1}
\newcommand{\modelsNoConvergeAROneHeavyBase}{0}
\newcommand{\percImputedPops}{17}
\newcommand{\percImputedPoints}{0.7}
\newcommand{\nuCoefPopN}{606}
\newcommand{\AvesRangePerc}{4--8}
\newcommand{\InsectaRangePerc}{2--3}
\newcommand{\MammaliaRangePerc}{4--6}
\newcommand{\OsteichthyesRangePerc}{0}
\newcommand{\overallMinPerc}{3}
\newcommand{\overallMaxPerc}{5}
\newcommand{\overallBasePerc}{4}
\newcommand{\NPops}{609}
\newcommand{\NOrders}{39}
\newcommand{\NClasses}{7}
\newcommand{\interpPointsPerc}{1}
\newcommand{\nBSUp}{8}
\newcommand{\nBSDown}{51}
\newcommand{\ratioBSDownToUp}{6.4}
\newcommand{\percBSDown}{86}
\newcommand{\crashUnderRange}{1.1--2}
\newcommand{\crashUnderMedian}{1.3}
\newcommand{\probDensSkewedForHeavyPops}{86}
\newcommand{\percNormPopsNotSkewed}{88}

@@

We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
The Gompertz model represents population growth as a linear function in
log space (here and throughout we use log
to refer to the natural logarithm).
If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then
\begin{align}
x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
  \epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
\end{align}
The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
= 1\). The model is density independent if \(b = 1\), maximally density
dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
allow it to be drawn from a Student-$t$ distribution with scale parameter
\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
which the process deviations have heavy tails and are therefore evidence of
black-swan events (Fig.~1A, B).

@@

We fit our models with Stan 2.12.1 \cite{hoffman2014, stan-manual2015, carpenter2016}, and R
3.3.1 \cite{r2015}. We began with four chains and 2,000 iterations, discarding
the first 1,000 as warm up. If \(\hat{R}\) (the
potential scale reduction factor---a measure of chain convergence) was greater
than \(1.05\) for any parameter, or if the minimum effective sample size,
\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
for any parameter was less than \(200\), we doubled both the total iterations
and warm up period and sampled from the model again.
In the majority of cases our minimum thresholds were
greatly exceeded. We continued this procedure up to 8,000 iterations
by which point all chains had sufficiently converged.
% The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
% requires fewer iterations to obtain equivalent effective sample sizes than the
% Gibbs or Metropolis--Hastings algorithms \cite{stan-manual2015}.
% \cite{monnahan2016}

>>>>

% \input{../analysis/values} % R output
\newcommand{\basePriorMean}{102}
\newcommand{\basePriorMedian}{71}
\newcommand{\basePriorProbHeavy}{7.7}
\newcommand{\medianTimeSteps}{26}
\newcommand{\meanTimeSteps}{30.2}
\newcommand{\minTimeSteps}{20}
\newcommand{\maxTimeSteps}{117}
\newcommand{\birdN}{191}
\newcommand{\insectsN}{182}
\newcommand{\mammalsN}{125}
\newcommand{\fishN}{108}
\newcommand{\birdNH}{14}
\newcommand{\insectsNH}{5}
\newcommand{\mammalsNH}{6}
\newcommand{\fishNH}{0}
\newcommand{\birdPH}{7}
\newcommand{\insectsPH}{3}
\newcommand{\mammalsPH}{5}
\newcommand{\fishPH}{0}
\newcommand{\propSkewedAves}{0.21}
\newcommand{\propSkewedInsecta}{0.18}
\newcommand{\propSkewedMammalia}{0.31}
\newcommand{\propSkewedOsteichthyes}{0.21}
\newcommand{\NOrdersHeavy}{15}
\newcommand{\POrdersHeavy}{38}
\newcommand{\baseFiftyObsFiftySwitch}{8}
\newcommand{\baseSeventyFiveObsFiftySwitch}{2}
\newcommand{\totalHeavyFifty}{26}
\newcommand{\totalHeavySeventyFive}{17}
\newcommand{\baseFiftyObsFiftySwitchPerc}{31}
\newcommand{\baseSeventyFiveObsFiftySwitchPerc}{8}
\newcommand{\baseNuTenObsTenSwitch}{8}
\newcommand{\baseNuTen}{26}
\newcommand{\baseNuFiveObsTenSwitch}{2}
\newcommand{\pHeavyNThirty}{0.13}
\newcommand{\pHeavyNSixty}{0.20}
\newcommand{\pIncHeavyNThirtyNSixty}{1.5}
\newcommand{\obsErrorNuFivePerc}{75}
\newcommand{\modelsNoConvergeAROne}{1}
\newcommand{\modelsNoConvergeAROneHeavyBase}{0}
\newcommand{\percImputedPops}{17}
\newcommand{\percImputedPoints}{0.7}
\newcommand{\nuCoefPopN}{606}
\newcommand{\AvesRangePerc}{4--8}
\newcommand{\InsectaRangePerc}{2--3}
\newcommand{\MammaliaRangePerc}{4--6}
\newcommand{\OsteichthyesRangePerc}{0}
\newcommand{\overallMinPerc}{3}
\newcommand{\overallMaxPerc}{5}
\newcommand{\overallBasePerc}{4}
\newcommand{\NPops}{609}
\newcommand{\NOrders}{39}
\newcommand{\NClasses}{7}
\newcommand{\interpPointsPerc}{1}
\newcommand{\nBSUp}{8}
\newcommand{\nBSDown}{49}
\newcommand{\ratioBSDownToUp}{6.1}
\newcommand{\percBSDown}{86}
\newcommand{\crashUnderRange}{1.2--1.9}
\newcommand{\crashUnderMedian}{1.4}
\newcommand{\probDensSkewedForHeavyPops}{86}
\newcommand{\percNormPopsNotSkewed}{89}

@@

We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
The Gompertz model represents population growth as a linear function in
log space (here and throughout we use log
to refer to the natural logarithm).
If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then
\begin{align}
x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
  \epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
\end{align}
The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
= 1\). The model is density independent if \(b = 1\), maximally density
dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
allow it to be drawn from a Student-$t$ distribution with scale parameter
\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
which the process deviations have heavy tails (i.e., heavier tails than the
lognormal distribution) and are therefore evidence of black-swan events
(Fig.~1A, B).

@@

We fit our models with Stan 2.14.1 \cite{hoffman2014, stan-manual2015, carpenter2016}, and R
3.3.1 \cite{r2015}. We began with four chains and 2,000 iterations, discarding
the first 1,000 as warm up. If \(\hat{R}\) (the
potential scale reduction factor---a measure of chain convergence) was greater
than \(1.05\) for any parameter, or if the minimum effective sample size,
\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
for any parameter was less than \(200\), we doubled both the total iterations
and warm up period and sampled from the model again.
In the majority of cases our minimum thresholds were
greatly exceeded. We continued this procedure up to 8,000 iterations
by which point all chains had sufficiently converged.
% The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
% requires fewer iterations to obtain equivalent effective sample sizes than the
% Gibbs or Metropolis--Hastings algorithms \cite{stan-manual2015}.
% \cite{monnahan2016}

### 2017-01-13 text/supp.tex
Resubmit

Since Stan only needs the probability density up to an additive
constant \cite{stan-manual2014}, we
can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
probability density for the skewed Student-$t$ distribution implemented in
Stan\footnote{Credit to Stan developer Bob Carpenter:
\url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
looks like the following:

@@

>>>>

Since Stan only needs the probability density up to an additive
constant \cite{stan-manual2014}, we
can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
probability density for the skewed Student-$t$ distribution implemented in
Stan\footnote{Credit to Stan developer Bob Carpenter:\\
\url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
looks like the following:

@@

In our skew-Student-$t$ models, populations tended to have downward-skewed
process noise (Figs 4B, \ref{fig:skew-nu}). The proportions of populations with
Pr($\gamma < 1$) $> 0.95$
(i.e.\ strong evidence of downward-skewed process noise)
were
\propSkewedAves\ (birds), \propSkewedInsecta\ (insects),
\propSkewedMammalia\ (mammals), and \propSkewedOsteichthyes\ (fishes).
